---
title: "HW #13 Computer Intensive Statistics in Ecology"
author: "Ä¬¥D©¶ Eric Su"
date: "2017/5/23"
output: html_document
---
[Click here for other works of the author on RPubs](http://rpubs.com/prorichter)

There are many functions that try to describe growth, in this assignment we would model our data using the von Bertalanffy and Gompertz growth function.

The von Bertalanffy growth function assumes a linear decrease in growth rate. It is defined as $L(t) = L_{\infty}(1 - e^{- K(t - t_0)})$ where $L_{\infty}$ is the hypothesized (mean) maximum length, $K$ is growth rate, $t$ is current age and $t_0$ is the hypothesized age when body length is 0. 

The Gompertz growth function assumes a exponential decrease in growth rate. It is defined as $L(t) = L_{\infty} * e^{-e^{- K_2(t - \frac{ln(\lambda) - ln(K_2))}{K_2})})}$ where $L_{\infty}$ is the hypothesized (mean) maximum length, $K_2$ is the rate of exponential decrease of the relative growth rate with age, $t$ is current age and $\lambda$ is the theoretical initial relative growth rate at zero age.

We would try to model body length of fish using regression models with multiplicative log-normal errors for both functions described above: $L(t) = Function * e^{\epsilon}$ with $\epsilon$ ~ $N(0, \sigma^2)$ and estimate the parameters using maximum likelihood estimators (MLE). Afterwards, we estimate the 95% confidence interval for model parameters and compare the goodness of fit using AICc and also find an average model using AIC weights.

### Load package
```{r}
library(knitr)
library(ggplot2)
```

### Load data
```{r}
data = read.csv("DATA_atlantic croaker.csv")

#keep only female data
data = subset(data, Sex == "F")[, -3]

#show a part of data
kable(head(data), row.names = F, caption = "Age and body length data of female Atlantic Croaker (partial)")
```

## Q1. Estimate the model parameters of two different growth functions (VBGF; Gompertz) using the method maximum likelihood for the length-at-age for female Atlantic Croaker (Micropogonias undulatus), considering a multiplicative error model with log-normal distribution

### Explore the relationship between body length and age
```{r}
#prepare data for plotting
dataset = data
colnames(dataset) <- c("Age", "Body length")

#draw graph
ggplot(dataset, aes(x = Age, y = `Body length`)) +
    geom_point()
```

Since age is theoretically a continuous variable, let's add a bit of jittering(noise) to simulate values for non-integer ages.
```{r}
ggplot(dataset, aes(x = Age, y = `Body length`)) +
    geom_jitter()
```

It does seem that growth rate declines as age increases.

### The von Bertalanffy growth function
Define function `VBGF` that calculates estimated length using the von Bertalanffy growth function.
```{r}
VBGF <-function(x, Linf = theta[1], K = theta[2], t0 = theta[3], theta = c(Linf, K, t0)){
    y = Linf * (1 - exp(- K * (x - t0)))
    return(y)
}
```

### The Gompertz growth function
Define function `Gomp` that calculates estimated length using the Gompertz growth function.
```{r}
Gomp <-function(x, Linf = theta[1], K2 = theta[2], lambda = theta[3], theta = c(Linf, K2, lambda)){
    y = Linf * exp(- exp(- K2 * (x - (log(lambda) - log(K2)) / K2)))
    return(y)
}
```

### Log-likelihood function for the regression
Define function `lognormal_like` which calculates the log-likelihood based on a multiplicative model with log-normal error
```{r}
lognormal_like = function(x, y, FUN, Out, ...){
    
    FUN = match.fun(FUN)
    like = numeric(length(y))
    NLL = numeric(length(y))
    ypred = FUN(x, ...)
    dev2 =(log(y) - log(ypred)) ^ 2
    sigma = sqrt(mean(dev2))
    
    for (i in 1:length(y)){
        like[i] = (1 / (x[i] * sqrt(2 * pi) * sigma)) * exp(-dev2[i] / (2 * sigma ^ 2))
        NLL[i] = -log(like[i])
    }
    
    #show each iteration if Out is set as TRUE
    if(Out == T) cat("NLL=", sum(NLL), " Linf=", Linf, " K=", K, " t0=", t0, "\n", sep = "")
    
    #calculate various informations about the log-likelihood
        Outs <- NULL
        Outs$Pred <- ypred
        Outs$Length <- y
        Outs$Dev2 <- dev2
        Outs$sigma <- sigma
        Outs$Like <- like
        Outs$NLL <- NLL 
        Outs$Obj <- sum(NLL)
        
    return(Outs)
}
```

### Define objective function: sum of negative log-likelihood
```{r}
obj_function = function(theta, ...){ 
    Outs <- lognormal_like(theta = theta, ...)
    obj <- Outs$Obj
    return(obj)
}
```

### Find the MLE of parameters
Using optimization method, we try to find values of the three parameters ($L_{\infty}, K, t_0$) that maximizes the likelihood function (=minimize negative log-likelihood function) for our regression model.
```{r warning=FALSE}
model_VBGF = optim(c(max(data[, 2]), 0.7, 0), obj_function, FUN = VBGF, x = data[, 1], y = data[ ,2], Out = F, hessian = T)
model_Gomp = optim(c(max(data[, 2]), 0.3, 0.2), obj_function, FUN = Gomp, x = data[ ,1], y = data[ ,2], Out = F, hessian = T)
```

### Display model results
```{r}
param_VBGF <- model_VBGF$par
param_Gomp <- model_Gomp$par
sd_VBGF <- sqrt(diag(solve(model_VBGF$hessian)))
sd_Gomp <- sqrt(diag(solve(model_Gomp$hessian)))
model_result_VBGF <- cbind(param_VBGF, sd_VBGF)
model_result_Gomp <- cbind(param_Gomp, sd_Gomp)
rownames(model_result_VBGF) <- c("L inf", "K", "t0")
rownames(model_result_Gomp) <- c("L inf", "K2", "Lambda")

kable(model_result_VBGF, col.names = c("Estimated parameters", "Standard deviation"), caption = "Model using von Bertalanffy growth function", digits = 3)
```

Regression model using the von Bertalanffy growth function:
$L(t) = 432.38(1 - e^{- 0.24(t + 1.98)})$
```{r}
kable(model_result_Gomp, col.names = c("Estimated parameters", "Standard deviation"), caption = "Model using Gompertz growth function", digits = 3)
```

Regression model using the Gompertz growth function:
$L(t) = 420.39 * e^{-e^{- 0.32(t - \frac{ln(0.28) - ln(0.32))}{0.32})})}$

## Q2. Calculate the 95% confidence intervals for each parameter of the growth functions by using the likelihood profile method

### Calculate negative log-likelihood for both models
```{r}
VBGF_like = lognormal_like(data[, 1], data[, 2], VBGF, theta = param_VBGF, Out = F)$Obj
Gomp_like = lognormal_like(data[, 1], data[, 2], Gomp, theta = param_Gomp, Out = F)$Obj
```

### Find 95% confidence interval for von Bertalanffy model parameters using the likelihood profile method
CI for $L_{\infty}$
```{r}
obj_fun = function(theta, ...){ 
    Outs <- lognormal_like(K = theta[1], t0 = theta[2], ...)
    obj <- Outs$Obj
    return(obj)
}

Linf_like = numeric()
Linf_range = seq(from = param_VBGF[1] - 40, to = param_VBGF[1] + 100, length.out = 100)
for(i in 1:100){
        Linf_est = optim(param_VBGF[c(2, 3)], obj_fun, FUN = VBGF, Linf = Linf_range[i], x = data[, 1], y = data[ ,2], Out = F)$par
        Linf_like[i] = 2 * (- VBGF_like + lognormal_like(data[, 1], data[, 2], VBGF, Linf =  Linf_range[i], K = Linf_est[1], t0 = Linf_est[2], Out = F)$Obj)
}

Linf_accept = which(Linf_like < qchisq(0.95, 1))
Linf_ci = Linf_range[c(min(Linf_accept), max(Linf_accept))]

```

CI for $K$
```{r warning=FALSE}
obj_fun = function(theta, ...){ 
    Outs <- lognormal_like(Linf = theta[1], t0 = theta[2], ...)
    obj <- Outs$Obj
    return(obj)
}

K_like = numeric()
K_range = seq(from = param_VBGF[2] - 0.15, to = param_VBGF[2] + 0.3, length.out = 100)
for(i in 1:100){
        K_est = optim(param_VBGF[c(1, 3)], obj_fun, FUN = VBGF, K = K_range[i], x = data[, 1], y = data[ ,2], Out = F)$par
        K_like[i] = 2 * (- VBGF_like + lognormal_like(data[, 1], data[, 2], VBGF, K =  K_range[i], Linf = K_est[1], t0 = K_est[2], Out = F)$Obj)
}

K_accept = which(K_like < qchisq(0.95, 1))
K_ci = K_range[c(min(K_accept), max(K_accept))]
```

CI for $t_0$
```{r warning=FALSE}
obj_fun = function(theta, ...){ 
    Outs <- lognormal_like(Linf = theta[1], K = theta[2], ...)
    obj <- Outs$Obj
    return(obj)
}

t0_like = numeric()
t0_range = seq(from = param_VBGF[3] - 2.04, to = param_VBGF[3] + 2.04, length.out = 100)
for(i in 1:100){
        t0_est = optim(param_VBGF[c(1, 2)], obj_fun, FUN = VBGF, t0 = t0_range[i], x = data[, 1], y = data[ ,2], Out = F)$par
        t0_like[i] = 2 * (- VBGF_like + lognormal_like(data[, 1], data[, 2], VBGF, t0 =  t0_range[i], Linf = t0_est[1], K = t0_est[2], Out = F)$Obj)
}

t0_accept = which(t0_like < qchisq(0.95, 1))
t0_ci = t0_range[c(min(t0_accept), max(t0_accept))]
```

### Display von Bertalanffy model results with confidence interval
```{r}
model_result_VBGF = data.frame(model_result_VBGF, t(matrix(c(Linf_ci, K_ci, t0_ci), 2, 3)))

kable(model_result_VBGF, col.names = c("Estimated parameters", "Standard deviation", "95% CI lower bound", "95% CI upper bound"), caption = "Model using von Bertalanffy growth function", digits = 3)
```

### Find 95% confidence interval for Gompertz model parameters using the likelihood profile method
CI for $L_{\infty}$
```{r}
obj_fun = function(theta, ...){ 
    Outs <- lognormal_like(K2 = theta[1], lambda = theta[2], ...)
    obj <- Outs$Obj
    return(obj)
}

LGomp_like = numeric()
LGomp_range = seq(from = param_Gomp[1] - 40, to = param_Gomp[1] + 80, length.out = 100)
for(i in 1:100){
        LGomp_est = optim(param_Gomp[c(2, 3)], obj_fun, FUN = Gomp, Linf = LGomp_range[i], x = data[, 1], y = data[ ,2], Out = F)$par
        LGomp_like[i] = 2 * (- Gomp_like + lognormal_like(data[, 1], data[, 2], Gomp, Linf =  LGomp_range[i], K2 = LGomp_est[1], lambda = LGomp_est[2], Out = F)$Obj)
}

LGomp_accept = which(LGomp_like < qchisq(0.95, 1))
LGomp_ci = LGomp_range[c(min(LGomp_accept), max(LGomp_accept))]

```

CI for $K_2$
```{r warning=FALSE}
obj_fun = function(theta, ...){ 
    Outs <- lognormal_like(Linf = theta[1], lambda = theta[2], ...)
    obj <- Outs$Obj
    return(obj)
}

K2_like = numeric()
K2_range = seq(from = param_Gomp[2] - 0.20, to = param_Gomp[2] + 0.3, length.out = 100)
for(i in 1:100){
        K2_est = optim(param_Gomp[c(1, 3)], obj_fun, FUN = Gomp, K2 = K2_range[i], x = data[, 1], y = data[ ,2], Out = F)$par
        K2_like[i] = 2 * (- Gomp_like + lognormal_like(data[, 1], data[, 2], Gomp, K2 =  K2_range[i], Linf = K2_est[1], lambda = K2_est[2], Out = F)$Obj)
}

K2_accept = which(K2_like < qchisq(0.95, 1))
K2_ci = K2_range[c(min(K2_accept), max(K2_accept))]
```

CI for $\lambda$
```{r warning=FALSE}
obj_fun = function(theta, ...){ 
    Outs <- lognormal_like(Linf = theta[1], K2 = theta[2], ...)
    obj <- Outs$Obj
    return(obj)
}

lambda_like = numeric()
lambda_range = seq(from = param_Gomp[3] - 0.15, to = param_Gomp[3] + 0.50, length.out = 100)
for(i in 1:100){
        lambda_est = optim(param_Gomp[c(1, 2)], obj_fun, FUN = Gomp, lambda = lambda_range[i], x = data[, 1], y = data[ ,2], Out = F)$par
        lambda_like[i] = 2 * (- Gomp_like + lognormal_like(data[, 1], data[, 2], Gomp, lambda =  lambda_range[i], Linf = lambda_est[1], K2 = lambda_est[2], Out = F)$Obj)
}

lambda_accept = which(lambda_like < qchisq(0.95, 1))
lambda_ci = lambda_range[c(min(lambda_accept), max(lambda_accept))]
```

### Display Gompertz model results with confidence interval
```{r}
model_result_Gomp = data.frame(model_result_Gomp, t(matrix(c(LGomp_ci, K2_ci, lambda_ci), 2, 3)))

kable(model_result_Gomp, col.names = c("Estimated parameters", "Standard deviation", "95% CI lower bound", "95% CI upper bound"), caption = "Model using Gompertz growth function", digits = 3)
```

## Q3. Conduct model selection based on the AICc and quantify the plausibility of each model by using ¡§Akaike weight ($w_i$)¡¨
```{r}
#calculate AICc for both models
aicc_VBGF = 2 * VBGF_like + 2 * 3 + (2 * 3 * 4) / (204 - 3 - 1)
aicc_Gomp = 2 * Gomp_like + 2 * 3 + (2 * 3 * 4) / (204 - 3 - 1)

#find the best model with the lower AICc
min_aicc = min(aicc_VBGF, aicc_Gomp)

#calculate weights based on AICc
aicc_change = c(aicc_VBGF, aicc_Gomp) - min_aicc
weight = exp(-0.5 * aicc_change) / sum(exp(-0.5 * aicc_change))

#show results of AICc and weights
m_criterion = cbind(c(aicc_VBGF, aicc_Gomp), aicc_change, weight)
rownames(m_criterion) <- c("von Bertalanffy", "Gompertz")
colnames(m_criterion) <- c("AICc", "AICc change", "Weight")
kable(m_criterion)
```

The model using the von Bertalanffy growth function is the better model based on AICc, but the difference is extremely small.

## Q4. Estimate the average model based on $w_i$ and plot all the growth curves together

### Average model
Make predictions using the average model based on Akaike weight
```{r}
pred_VBGF = VBGF(seq(from = min(data[ ,1]), to = max(data[, 1]), length.out = 100), theta = param_VBGF)
pred_Gomp = Gomp(seq(from = min(data[ ,1]), to = max(data[, 1]), length.out = 100), theta = param_Gomp)
pred_ave = weight[1] * pred_VBGF + weight[2] * pred_Gomp
```

### Visualize estimations of regression model
```{r}
#prepare data for plotting
dataset = data.frame(rep(seq(from = min(data[ ,1]), to = max(data[, 1]), length.out = 100), 3), c(pred_VBGF, pred_Gomp, pred_ave), c(rep("VBGF", 100), rep("Gompertz",100), rep("Average", 100)))
colnames(dataset) <- c("Age", "Esimated length", "Model")

#draw graph
p <- ggplot() +
    geom_line(aes(x = dataset$Age, y = dataset$`Esimated length`, col = dataset$Model)) +
    scale_color_discrete(name = "Model") +
    labs(title = "Regression model using von Bertalanffy and Gompertz growth function", x = "Age", y = "Fish length")
p + geom_point(aes(x = data[, 1], y = data[, 2]))

#graph with jittering
p + geom_jitter(aes(x = data[, 1], y = data[, 2]))
```

We can see that all models make very similar estimations, and our model does fit the data well.