---
title: "HW #2 Computer Intensive Statistics in Ecology"
author: "Ä¬¥D©¶ Eric Su"
date: "2017/3/11"
output: html_document
---
[Click here for other works of the author on RPubs](http://rpubs.com/prorichter)

## 1a. Generate 10000 random numbers from Gaussian distribution with mean=20 and variance =10, and plot the distribution.

### function to generate random numbers from a normal distribution
Define function `r.norm` to generate random numbers from a normal distribution using samples from a standard normal distribution.
```{r}
r.norm <- function(num, mu, v){
    
    #check for invalid values
    if(num%%1 != 0 | num < 0) stop('The number of samples to be selected must be a positive integer')
    if(v <= 0) stop('Variance must be positive')
    
    #generate samples from a standard normal distribution and convert them 
    #according to the mean and variance given
    norm.sample = sqrt(v) * rnorm(num) + mu
    return(norm.sample)
}
```

### Generate and plot random samples from $N(20, 10)$
```{r}
#plot the distribution of the normal sample with a fitted normal curve
norm.samp = r.norm(num = 10000, mu = 20, v = 10)
hist(norm.samp, breaks = 30,col = "gray", main = "Histogram of the normal sample", xlab = "Value", prob = TRUE)
curve(dnorm(x, mean = mean(norm.samp), sd = sd(norm.samp)), col = "darkblue", lwd = 2, add = TRUE, yaxt = "n")
```

## 1b. Generate 10000 random numbers from Binomial distribution with p=0.5 and n=40, and plot the distribution. Compare the distribution of 1a and 1b, what do you find? (hint: you can use ¡§rand¡¨ and ¡§randn¡¨ in matlab. Or, ¡§runif¡¨ and ¡§rnorm¡¨ in R)

### function to generate random numbers from a binomial distribution
Define function `r.bin` to generate random numbers from a binomial distribution using samples from uniform distribution.
```{r}
r.bin <- function(num, p, n){
    
    #check for invalid values
    if(num%%1 != 0 | num <= 0) stop('The number of samples to be selected must be a positive integer')
    if(p < 0 | p > 1) stop('Parameter p must be between 0 and 1')
    if(n%%1 != 0 | n <= 0) stop('Parameter n must be a positive integer')
    
    #generate random numbers from a uniform distribution
    unif.sample = matrix(runif(n * num), n, num)
    
    #transform the samples into Bernoulli samples
    Bern.sample = apply(unif.sample, c(1, 2), function(x) if(x >= p){x = 0} else{x = 1})
    
    #sum Bernoulli samples into binomial samples
    bin.sample = colSums(Bern.sample)
    
    return(bin.sample)
}
```

### Generate and plot random samples from $Binom(0.5, 40)$
```{r}
#plot the distribution of the binomial sample with a fitted normal curve
bin.samp = r.bin(num = 10000, p = 0.5, n = 40)
hist(bin.samp, breaks = seq(min(bin.samp) - 0.5, max(bin.samp) + 0.5, by = 1),col = "gray", main = "Histogram of the binomial sample", xlab = "Value", prob = TRUE)
curve(dnorm(x, mean = mean(bin.samp), sd = sd(bin.samp)), col = "darkblue", lwd = 2, add = TRUE, yaxt = "n")
```

It is apperant that the binomial sample is also very similar to a normal distribution.

## 2. Make a program that can select our candidates for presentation next week. This program should select randomly but avoid selecting the numbers that had been selected before.

### function to randomly select samples from a population without replacement
```{r}
r.choose <- function(select, total){
    
    #check for invalid values
    if(select%%1 != 0 | select < 0) stop('The number of samples to be selected must be a positive integer')
    if(total%%1 != 0 | total <= 0) stop('The total number of population must be a positive integer')
    if(select > total) stop('The number of samples to be selected must not exceed the population size')
    
    #seperate situations which the number of people to be selected is more than
    #50% of the total to improve efficiency
    if(select / total <= 0.5 | select == total){
        #generate sample
        choose = ceiling(runif(select, 0, total))
        
        #replace duplicated numbers in the sample until all numbers are unique
        while(length(choose[duplicated(choose)]) > 0){
            choose[duplicated(choose)] = ceiling(runif(length(choose[duplicated(choose)]), 0, total))
        }
    }
    
    else{
        #generate sample for numbers not to be included
        choose = ceiling(runif(total-select, 0, total))
        
        #replace duplicated numbers in the sample until all numbers are unique
        while(length(choose[duplicated(choose)])>0){
            choose[duplicated(choose)] = ceiling(runif(length(choose[duplicated(choose)]), 0, total))
        }
        
        #remove numbers not to be included to get final sample
        choose = c(1:total)[-choose]
    }
    
    #return results
    return(sort(choose))
}
```

### Use the self defined function `r.choose` to select 2 random samples without replacement
```{r}
#select 2 people from 18
r.choose(select = 2, total = 18)
```