---
title: "HW #3 Computer Intensive Statistics in Ecology"
author: "Ä¬¥D©¶ Eric Su"
date: "2017/3/10"
output:
  html_document:
    toc: true
    toc_float: true
---
[Click here for other works of the author on RPubs](http://rpubs.com/prorichter)

## Import data
Import data set `enviANDdensity` and extract density data of fish and copepod from the data set
```{r}
library(knitr)
enviANDdensity <- read.table("enviANDdensity.txt", header = T)
fish_density=as.matrix(enviANDdensity[,11])
copepod_density=as.matrix(enviANDdensity[,12])
kable(head(cbind(fish_density,copepod_density)),digits=2,col.names = c("Fish density","Copepod density"),align="l",caption="Fish and copepod density data")
```

## 1. Compute the *mean* and *SE(mean)* for the fish and copepod densities respectively, using both normal theory and non-parametric bootstrap. Plot the histogram of bootstrapped means with bootstrap 1000 times.

### Function to generate bootstrap samples
Create function `boot.samp` to generate bootstrap samples for a given sample and a statistic of interest
```{r}
boot.samp <- function(samp,n,stat){
    FUN=match.fun(stat)
    
    #create and randomize n sets of bootstrap samples
    size=length(samp)
    shuffle=rep(samp,n)[order(runif(size*n))]
    
    #split bootstrap samples and calculate the statistic of interest for each set
    bt=apply(matrix(shuffle,size,n),2,FUN)
    
    #return bootstrap samples
    return(bt)
}
```

### Function to report bootstrap results
Create function `boot.result` to report the estimated statistic and its standard error.
```{r}
boot.result <- function(bt.samp,origin.mean=NULL){
    
    #calculate the estimated parameter and SE for bootstrap
    result=t(as.matrix(cbind(mean(bt.samp),sd(bt.samp))))
    
    # If mean is the statistic of interest, calculate SE using normal theory
    if(!is.null(origin.mean)){
        origin=t(as.matrix(cbind(mean(origin.mean),sd(origin.mean)/sqrt(length(origin.mean)))))
        result=cbind(origin,result)
        colnames(result) <- c("Normal theory","Bootstrap")
        rownames(result) <- c("Estimated parameter","Estimated standard error")
    }
    
    else{
        colnames(result) <- c("Bootstrap")
        rownames(result) <- c("Estimated parameter","Estimated standard error")
    }
    return(result)
}
```

### Function to plot a histogram for bootstrap samples
Create function `boot.plot` to plot the bootstrap samples with a histogram
```{r}
boot.plot <- function(bt.samp,p.title="Bootstrap histogram",hist.break=50){
    
    #plot the histogram of bootstrap samples along with a smoothed density curve and     95% confidence interval
    hist(bt.samp,breaks=hist.break,col="gray",main=p.title,prob=T,xlab=NULL)
    lines(density(bt.samp),col="blue",lwd=2)
    abline(v=sort(bt.samp)[length(bt.samp)*c(0.025, 0.975)],col="red",lwd=2)
    legend("topright",lty=c(1,1),legend=c("Smoothed density","95% CI"),col=c("blue","red"),lwd=2)
}
```

### Generate bootstrap samples for the mean of fish and copepod density 
Bootstrap samples are generated using the function created above. Sample size is set to 1000.
```{r}
fish_mean_bt=boot.samp(fish_density,1000,mean)
copepod_mean_bt=boot.samp(copepod_density,1000,mean)
```

### Bootstrap results for the mean of fish density
```{r}
boot.fish.mean=boot.result(fish_mean_bt,fish_density)
kable(boot.fish.mean,digits=2,caption="Comparison of normal theory v.s bootstrap for fish density mean",align="l")
boot.plot(fish_mean_bt,"Bootstrap samples of fish density mean")
```

### Bootstrap results for the mean of copepod density
```{r}
boot.copepod.mean=boot.result(copepod_mean_bt,copepod_density)
kable(boot.copepod.mean,digits=2,caption="Comparison of normal theory v.s bootstrap for copepod density mean",align="l")
boot.plot(copepod_mean_bt,"Bootstrap samples of copepod density mean")
```

## 2. Compute the *median* and bootstrapped *SE(median)* for the fish and copepod densities.Plot the histogram of bootstrapped medians with bootstrap 1000 times.

### Generate bootstrap samples for the median of fish and copepod density
Bootstrap sampls are generated using the function created above. Sample size is set to 1000.
```{r}
fish_median_bt=boot.samp(fish_density,1000,median)
copepod_median_bt=boot.samp(copepod_density,1000,median)
```

### Bootstrap results for the median of fish density
```{r}
boot.fish.median=boot.result(fish_median_bt)
kable(boot.fish.median,digits=2,caption="Bootstrap result for fish density median",align="l")
boot.plot(fish_median_bt,"Bootstrap samples of fish density median")
```

### Bootstrap results for the median of copepod density
```{r}
boot.copepod.median=boot.result(copepod_median_bt)
kable(boot.copepod.median,digits=2,caption="Bootstrap result for copepod density median",align="l")
boot.plot(copepod_median_bt,"Bootstrap samples of copepod density median")
```

## 3a. Plot fish (dependent) v.s copepod (independent) and the regression line.
```{r}
# set dependent variable
Y=fish_density

# set intercept and independent variable
X=cbind(rep(1,length(copepod_density)),copepod_density)

# compute coefficients
B = solve(t(X) %*% X) %*% (t(X) %*% Y)

# plot fish  v.s copepod
plot(copepod_density,fish_density,xlab="Copepod density",ylab="Fish density",pch=16,col="blue")

# add the regression line
x=seq(min(copepod_density),max(copepod_density),length=100)
lines(x,B[1]+B[2]*x,col="red",lwd=2)
legend("bottomright",pch=c(16,NA),lty=c(0,1),legend=c("Actual","Predicted"),col=c("blue","red"),lwd=2)
```

## 3b. Compute the regression coefficients for the linear model $Fish = \beta_0+\beta_1 \times Copepod$ and bootstrapped *SE(£]0)* and *SE(£]1)*. Plot the histogram of bootstrapped *£]0* and *£]1* with bootstrap 1000 times.

### Function to generate bootstrap samples for coefficients in a model
Create function `lm.boot` to generate bootstrap samples for coefficients in a linear regression model for given dependent and independent variables
```{r}
lm.boot <- function(dv,iv,n){
    
    B.sample=numeric()
    
    for(i in 1:n){
        
        #generate bootstrap samplea for coefficients in a linear regression model
        size=length(dv)
        samp=as.matrix(cbind(dv,iv))[ceiling(runif(size,0,size)),]
        Y=samp[,1]
        X=cbind(rep(1,length(samp[,1])),samp[,-1])
        beta = solve(t(X) %*% X) %*% (t(X) %*% Y)
        B.sample=append(B.sample,beta)
    }
    #return bootstrap samples for the coefficients
    return(t(matrix(B.sample,dim(samp)[2],n)))
}
```

### Bootstrap for coefficients in linear model $Fish = \beta_0+\beta_1 \times Copepod$
Conduct bootstrap on coefficients $B_0$, $B_1$ and report results using the `lm.boot` and `boot.result` functions created above
```{r}
lm.bt=lm.boot(fish_density,copepod_density,1000)
```

### Report bootstrap results for coefficient B0
```{r}
result.lm.B0=boot.result(lm.bt[,1])
kable(result.lm.B0,digits=2,caption="Bootstrap result of coefficient B0",align="l")
boot.plot(lm.bt[,1],"Histogram for bootstrap samples of coefficient B0")
```

### Report bootstrap results for coefficient B1
```{r}
result.lm.B1=boot.result(lm.bt[,2])
kable(result.lm.B1,digits=2,caption="Bootstrap result of coefficient B1",align="l")
boot.plot(lm.bt[,2],"Histogram for bootstrap samples of coefficient B1")
```