---
title: "HW #1 Computer Intensive Statistics in Ecology"
author: "Ä¬¥D©¶ Eric Su"
date: "2017/3/2"
output: html_document
---
[Click here for other works of the author on RPubs](http://rpubs.com/prorichter)

### Load package
```{r}
library(knitr)
```

## 1. Calculate the copepod density for each species for each cruise-station

### Import data
```{r}
species <- read.table("copepod_composition.txt",header = T) #import copepod data
dens <- as.vector(read.table("cop_density.txt",header = T)[[1]]) # import density data
```

### Data manipulation
```{r}
#convert species frequency into percentage
species = species / 100

#calculate copepod density for each species for each cruise station
species.density = t(apply(species, 1, function(x) x * dens))

kable(t(head(species.density)), digits = 3, col.names = 1:6)

```

## 2. For each cruise-station, calculate the species richness (number of species) and Shannon diversity index

```{r}
#calculate number of species in each cruise station
species_number = apply(species, 2, function(x) length(x[x > 0]))
kable(species_number, col.names = "Number of species")

#calculate Shannon diversity index for each cruise station
Shannon_index = apply(species, 2, function(x) - sum(x * log(x), na.rm = T))
kable(Shannon_index, col.names = "Shannon diversity index", digits = 2)
```

## 3. Find dominant species (species >=2% of total composition in any cruise-station) and calculate the average density for the spring, summer, and winter cruise for each dominant species.

### Find dominant species in spring
```{r}
#find dominant species in each station during spring
p.dom = apply(species[, 1:10] >= 0.02, 2, function(x) which(x == TRUE))
p.dom.species = sort(unique(unlist(p.dom)))

#calculate average density for each spring dominant species during each season
p.dom.spring = apply(species.density[p.dom.species, 1:10 ], 1, mean)
p.dom.summer = apply(species.density[p.dom.species, 11:25], 1, mean)
p.dom.winter = apply(species.density[p.dom.species, 26:34], 1, mean)
p.dom.result = cbind(p.dom.spring, p.dom.summer, p.dom.winter)
rownames(p.dom.result) = p.dom.species

#show density of spring dominant species in each season
kable(p.dom.result, col.names = c("Spring", "Summer", "Winter"), caption = "Density of spring dominant species", digits = 3)
```

### Find dominant species in summer
```{r}
#find dominant species in each station during summer
s.dom=apply(species[,11:25]>=0.02,2,function(x) which(x==TRUE))
s.dom.species=sort(unique(unlist(s.dom)))

#calculate average density for each summer dominant species during each season
s.dom.spring=apply(species.density[s.dom.species,1:10],1,mean)
s.dom.summer=apply(species.density[s.dom.species,11:25],1,mean)
s.dom.winter=apply(species.density[s.dom.species,26:34],1,mean)
s.dom.result=cbind(s.dom.spring, s.dom.summer, s.dom.winter)
rownames(s.dom.result) = s.dom.species

#show density of summer dominant species in each season
kable(s.dom.result, col.names = c("Spring", "Summer", "Winter"), caption = "Density of summer dominant species", digits = 3)
```

### Find dominant species in Winter
```{r}
#find dominant species in each station during winter
w.dom=apply(species[,26:34]>=0.02,2,function(x) which(x==TRUE))
w.dom.species=sort(unique(unlist(w.dom)))

#calculate average density for each winter dominant species during each season
w.dom.spring=apply(species.density[w.dom.species,1:10],1,mean)
w.dom.summer=apply(species.density[w.dom.species,11:25],1,mean)
w.dom.winter=apply(species.density[w.dom.species,26:34],1,mean)
w.dom.result=cbind(w.dom.spring, w.dom.summer, w.dom.winter)
rownames(w.dom.result) = w.dom.species

#show density of winter dominant species in each season
kable(w.dom.result, col.names = c("Spring", "Summer", "Winter"), caption = "Density of winter dominant species", digits = 3)
```