---
title: "HW #14 Computer Intensive Statistics in Ecology"
author: "Ä¬¥D©¶ Eric Su"
date: "2017/5/25"
output: html_document
---
[Click here for other works of the author on RPubs](http://rpubs.com/prorichter)

The von Bertalanffy growth function is used to describe body length of fish as a function of its age. It is defined as $L(t) = L_{\infty}(1 - e^{- K(t - t_0)})$ where $L_{\infty}$ is the hypothesized (mean) maximum length, $K$ is growth rate, $t$ is current age and $t_0$ is the hypothesized age when body length is 0.

In this assignment, we try to model body length of fish using the following regression model: $L(t) = L_{\infty}(1 - e^{-K(t - t_0)}) * e^{\epsilon}$ with $\epsilon$ ~ $N(0, \sigma^2)$. Parameters will be estimated using the Bayesian grid search with priors $L_{\infty}$ ~ $U(40, 100)$ and $K$ ~ $U(0.1, 0.6)$.

## Q. Assume $t_0$=0 (two parameters). Please estimate the posterior distributions of parameters using the method Bayesian grid search for the average length-at-age for the female Pacific Hake, respectively, considering a multiplicative error model with log-normal distribution.

### Load package
```{r}
library(knitr)
library(ggplot2)
```

### Load data
```{r}
age = c(1, 2, 3.3, 4.3, 5.3, 6.3, 7.3, 8.3, 9.3, 10.3, 11.3, 12.3, 13.3)
length = c(15.4, 28.03, 41.18, 46.2, 48.23, 50.26, 51.82, 54.27, 56.98, 58.93, 59, 60.91, 61.83)

#show data
kable(cbind(age, length), col.names = c("Age (year)", "Female mean length (cm)"))
```

### Explore the relationship between body length and age
```{r}
#prepare data for plotting
dataset = data.frame(age, length)
colnames(dataset) <- c("Age", "Body length")

#draw graph
ggplot(dataset, aes(x = Age, y = `Body length`)) +
    geom_point()
```

It seems that growth rate declines as age increases and the von Bertalanffy growth function is appropriate.

### The von Bertalanffy growth function
Define function `VBGF` that calculates estimated length using the von Bertalanffy growth function, which would be used in our regression model. The parameter $t_0$ would be set to 0.
```{r}
VBGF <-function(x, Linf, K){
    y = Linf * (1 - exp(- K * (x - 0)))
    return(y)
}
```

### Log-likelihood function for the regression
Define function `lognormal_like` which calculates the total likelihood of the regression model. The parameter sigma would be set to 0.2.
```{r}
lognormal_like = function(Linf, K){
    
    like = numeric(length(length))
    LL = numeric(length(length))
    ypred = VBGF(age, Linf, K)
    dev2 =(log(length) - log(ypred)) ^ 2
    sigma = 0.2
    
    for (i in 1:length(length)){
        like[i] = (1 / (age[i] * sqrt(2 * pi) * sigma)) * exp(-dev2[i] / (2 * sigma ^ 2))
        LL[i] = log(like[i])
    }
    
    tot_like = exp(sum(LL))
    return(tot_like)
}
```

### Assign the prior distribution values in a grid
Prior distributions for our parameters are: $L_{\infty}$ ~ $U(40, 100)$ and $K$ ~ $U(0.1, 0.6)$
```{r}
Linf_grid = seq(40, 100, length.out = 1000)
K_grid = seq(0.1, 0.6, length.out = 1000)

# prior porbability
pLinf_grid = rep(1, length(Linf_grid))
pK_grid = rep(1, length(K_grid))
pTheta_matrix = expand.grid(Linf = pLinf_grid, K = pK_grid)

# expand the combination of grid 
Theta_matrix = expand.grid(Linf = Linf_grid, K = K_grid)
```

### Calculate posterior probabilities
```{r}
# negtive likelihood values of each grid
pDataGivenTheta = mapply(lognormal_like, Theta_matrix[,1], Theta_matrix[,2])

# Compute the posterior density
pData = sum(pDataGivenTheta * pTheta_matrix[, 1] * pTheta_matrix[, 2])
pThetaGivenData = (pDataGivenTheta * pTheta_matrix[, 1] * pTheta_matrix[, 2]) / pData

# arrange the results
result = cbind(Theta_matrix, pThetaGivenData)
```

### Plot posterior distributions of parameters $L_{\infty}$ and $K$
```{r}
par(mfrow=c(2,1),mar=c(4,4,2,2))

# calculate the marginal probability
marginal_Linf = with(result, tapply(pThetaGivenData, Linf, sum))
plot(Linf_grid, marginal_Linf, type = "h", lwd = 3, xlab = expression(L[infinity]), ylab = bquote(paste("p(", theta, "|D)")))

# calculate the marginal probability
marginal_K = with(result, tapply(pThetaGivenData, K, sum))
plot(K_grid, marginal_K, type = "h", lwd = 3, xlab = "K", ylab = bquote(paste("p(", theta, "|D)")))
```