---
title: 'HW #6 Computer Intensive Statistics in Ecology'
author: "Ä¬¥D©¶ Eric Su"
date: "2017/3/31"
output: html_document
---
[Click here for other works of the author on RPubs](http://rpubs.com/prorichter)

## Load packages 
Load `knitr` for better report quality with markdown and `ggplot2` for better plots
```{r}
library(knitr)
library(ggplot2)
```

***
## Q1. Compute the regression coefficients for $fish = \beta_0 + \beta_1 * copepod$ and use randomization method to generate null distribution of $\beta_1$ and test whether $\beta_1$ is significantly different from null (with randomization 5000 times). Report your *p*-value. 

### Import data
Import data set `enviANDdensity`, extract density data of fish and copepod from the data set and display partial data
```{r}
enviANDdensity <- read.table("enviANDdensity.txt", header = T)
fish_density = as.matrix(enviANDdensity[, 11])
copepod_density = as.matrix(enviANDdensity[, 12])

# display a part of the data
kable(head(cbind(fish_density, copepod_density)), digits = 2, col.names = c("Fish density", "Copepod density"), align = "l", caption = "Fish and copepod density data (partial)")
```

### Calculate estimated $\hat{\beta_0}$ & $\hat{\beta_1}$
```{r}
Y = fish_density
X = cbind(1, copepod_density)
beta = solve(t(X) %*% X) %*% (t(X) %*% Y)
```

### Generate samples for permutaion test under null hypothesis $H_0:\beta_1=0$
```{r}
resamp_n = 100000
size = length(fish_density)
beta_sample = numeric()
for (i in 1:resamp_n) {
    #create a lish of indices for randomization
    shuffle = order(runif(size))
    
    #define randomized dependent variable and independent variables and calculate coefficients
    Y = fish_density[shuffle]
    X = cbind(1, copepod_density)
    b = solve(t(X) %*% X) %*% (t(X) %*% Y)
    beta_sample = append(beta_sample, b)
}
beta_sample = t(matrix(beta_sample, 2, resamp_n))
```

### Plot the distribution of permuted $\hat{\beta_1}$ samples and report *p*-value for the test
```{r message = FALSE}
#calculate p-value for the two tailed test
p_value_q1 = sum(abs(beta_sample[, 2]) > abs(beta[2])) / resamp_n

#create density plot for the permuted samples and add vertical lines at the estimated value
lab_q1 = c(-beta[2], seq(min(beta_sample[, 2]), max(beta_sample[, 2]), length.out=5), beta[2])
p_q1 <- ggplot(as.data.frame(beta_sample[, 2]), aes(x = beta_sample[, 2])) +
        geom_density(fill = "#99CCFF", col = "#3399FF", alpha = 0.5) +
        geom_vline(xintercept = c(-beta[2], beta[2]), col = "red") +
        labs(title = expression(Distribution~of~permuted~hat(beta)[1]~samples~"for"~null~hypothesis), x = expression(Permuted~hat(beta)[1])) +
        annotate("text", x = beta[2], y = 10, label = "Esimated~hat(beta)[1]", hjust = 1.1, parse = T) +
        scale_x_continuous(breaks = lab_q1, labels = format(lab_q1, digits = 1))

#add shaded area to indicate significant values
d_q1 <- ggplot_build(p_q1)$data[[1]]
p_q1 + geom_area(data = subset(d_q1, x > beta[2]), aes(x = x, y = y), col = "red", fill = "red") +
    geom_area(data = subset(d_q1, x < -beta[2]), aes(x = x, y = y), col = "red",  fill = "red")
```

The *p*-value of our test on $H_0:\beta_1=0$ is `r format(p_value_q1, scientific=FALSE)`, therefore we `r if(p_value_q1 >= 0.05) "don't"` reject the null hypothesis and conclude that $\beta_1$ `r if(p_value_q1 < 0.05){"is"} else{"isn't"}` significantly different from 0.

***
## Q2. Randomization test whether significant difference exists between the density of Oncaea Venusta and Canthocalanus pauper. (Assume all data are independent and use all 34 stations.) Report your *p*-value.

### Import data
Import data `copepod_composition.txt`, `cop_density.txt` and `copepodSPlist.txt` for analysis
```{r}
#import copepod data
species <- read.table("copepod_composition.txt", header = T)

#import density data
dens <- as.vector(read.table("cop_density.txt", header = T)[[1]])

#import species name
species_name <- read.table("copepodSPlist.txt", sep = "\t")
```

### Extract Oncaea Venusta and Canthocalanus pauper data
```{r}
#convert species frequency into percentage
species = species / 100

#calculate copepod density for each species for each cruise station
species.density = t(apply(species, 1, function(x) x * dens))

#extract the density of Oncaea Venusta and Canthocalanus pauper
OV = species.density[grep("Oncaea venusta", species_name[, 1]), ]
CP = species.density[grep("Canthocalanus pauper", species_name[, 1]), ]

#show partial data
kable(head(cbind(OV, CP)), digits = 2, col.names = c("Oncaea Venusta", "Canthocalanus pauper"), align = "l", caption = "Oncaea Venusta and Canthocalanus pauper density data (partial)")
```

### Generate samples for permutation test for null hypothesis $H_0:\theta_1-\theta_2=0$
```{r}
diff = numeric(resamp_n)
for(i in 1:resamp_n){
    shuffle = c(rep(c(1, 2), 34))[order(runif(68))]
    data = cbind(c(OV, CP), shuffle)
    g1 = subset(data, shuffle == 1)
    g2 = subset(data, shuffle == 2)
    diff[i] = mean(g1) - mean(g2)
}
```

### Plot the distribution of permuted $\hat{\theta_1}-\hat{\theta_2}$ samples and report *p*-value for the test
```{r message = FALSE}
#calculate estimated difference between Oncaea Venusta and Canthocalanus pauper density
est_diff = mean(OV) - mean(CP)

#calculate p-value
p_value_q2 = sum(abs(diff) > abs(est_diff)) / resamp_n

#create density plot for the permuted samples and add vertical lines at the estimated value
lab_q2 = c(-est_diff, seq(min(diff), max(diff), length.out = 5), est_diff)
p_q2 <- ggplot(as.data.frame(diff), aes(x = diff)) +
        geom_density(fill = "#99CCFF", col = "#3399FF", alpha = 0.5) +
        geom_vline(xintercept = c(-est_diff, est_diff), col = "red") +
        labs(title = expression(Distribution~of~permuted~hat(theta)[1]-hat(theta)[2]~samples~"for"~null~hypothesis), x = expression(Permuted~hat(theta)[1]-hat(theta)[2])) +
        annotate("text", x = est_diff, y = 0.01, label = "Esimated~hat(theta)[1]-hat(theta)[2]", hjust = 1.1, parse = T) +
        scale_x_continuous(breaks = lab_q2, labels = format(lab_q2, digits = 2))

#add shaded area to indicate significant values
d_q2 <- ggplot_build(p_q2)$data[[1]]
p_q2 + geom_area(data = subset(d_q2, x > est_diff), aes(x = x, y = y), col = "red", fill = "red") +
    geom_area(data = subset(d_q2, x < -est_diff), aes(x = x, y = y), col = "red", fill = "red")
```

The *p*-value of our test on $H_0:\theta_1-\theta_2=0$ is `r format(p_value_q2, scientific=FALSE)`, therefore we `r if(p_value_q2 >= 0.05) "don't"` reject the null hypothesis and conclude that significant difference `r if(p_value_q2 < 0.05){"does"} else{"doesn't"}` exist between the density of Oncaea Venusta and Canthocalanus pauper.