---
title: "HW #12 Computer Intensive Statistics in Ecology"
author: "Ä¬¥D©¶ Eric Su"
date: "2017/5/12"
output: html_document
---
[Click here for other works of the author on RPubs](http://rpubs.com/prorichter)

The von Bertalanffy growth function is used to describe body length of fish as a function of its age. It is defined as $L(t) = L_{\infty}(1 - e^{- K(t - t_0)})$ where $L_{\infty}$ is the hypothesized (mean) maximum length, $K$ is growth rate, $t$ is current age and $t_0$ is the hypothesized age when body length is 0. 
In this assignment, we try to model body length of fish using the following regression model: $L(t) = L_{\infty}(1 - e^{-K(t - t_0)}) * e^{\epsilon}$ with $\epsilon$ ~ $N(0, \sigma^2)$ and estimate the parameters using maximum likelihood estimators (MLE).

### Load package
```{r}
library(knitr)
library(ggplot2)
```

### Load data
```{r}
age = c(1, 2, 3.3, 4.3, 5.3, 6.3, 7.3, 8.3, 9.3, 10.3, 11.3, 12.3, 13.3)
length_m = c(15.4, 26.93, 42.23, 44.59, 47.63, 49.67, 50.87, 52.3, 54.77, 56.43, 55.88)
length_f = c(15.4, 28.03, 41.18, 46.2, 48.23, 50.26, 51.82, 54.27, 56.98, 58.93, 59, 60.91, 61.83)

#show data
kable(cbind(age, length_f, c(length_m, rep("", 2))), col.names = c("Age (year)", "Female mean length (cm)", "Male mean length (cm)"))
```

### Explore the relationship between body length and age
```{r}
#prepare data for plotting
dataset = data.frame(c(age, age[-c(12:13)]), c(length_f, length_m), c(rep("Female", 13), rep("Male", 11)))
colnames(dataset) <- c("Age", "Body length", "Sex")

#draw graph
ggplot(dataset, aes(x = Age, y = `Body length`, col = Sex)) +
    geom_point() +
    scale_color_discrete(name = "Sex")
```

It seems that growth rate declines as age increases and the von Bertalanffy growth function is appropriate.

### The von Bertalanffy growth function
Define function `VBGF` that calculates estimated length using the von Bertalanffy growth function, which would be used in our regression model.
```{r}
VBGF <-function(x, Linf, K, t0){
    y = Linf * (1 - exp(- K * (x - t0)))
    return(y)
}
```

### Log-likelihood function for the regression
Define function `lognormal_like` which calculates the log-likelihood of 
```{r}
lognormal_like = function(x, y, Linf, K, t0, Out){
    
    like = numeric(length(y))
    NLL = numeric(length(y))
    ypred = VBGF(x, Linf, K, t0)
    dev2 =(log(y) - log(ypred)) ^ 2
    sigma = sqrt(mean(dev2))
    
    for (i in 1:length(y)){
        like[i] = (1 / (x[i] * sqrt(2 * pi) * sigma)) * exp(-dev2[i] / (2 * sigma ^ 2))
        NLL[i] = -log(like[i])
    }
    
    #show each iteration if Out is set as TRUE
    if(Out == T) cat("NLL=", sum(NLL), " Linf=", Linf, " K=", K, " t0=", t0, "\n", sep = "")
    
    #calculate various informations about the log-likelihood
        Outs <- NULL
        Outs$Pred <- ypred
        Outs$Length <- y
        Outs$Dev2 <- dev2
        Outs$sigma <- sigma
        Outs$Like <- like
        Outs$NLL <- NLL 
        Outs$Obj <- sum(NLL)
        
    return(Outs)
}
```

### Define objective function: sum of negative log-likelihood
```{r}
obj_function = function(theda, ...){ 
    Outs <- lognormal_like(Linf = theda[1], K = theda[2], t0 = theda[3], ...)
    obj <- Outs$Obj 
    return(obj)
}
```

### Find the MLE of parameters
Using optimization method, we try to find values of the three parameters ($L_{\infty}, K, t_0$) that maximizes the likelihood function (=minimize negative log-likelihood function) for our regression model. Female and male data would be fit seperately.
```{r warning=FALSE}
model_f = optim(c(100, 0.2, 0), obj_function, x = age, y = length_f, Out = F, hessian = T)
model_m = optim(c(100, 0.2, 0), obj_function, x = age[-c(12:13)], y = length_m, Out = F, hessian = T)
```

### Display model results
```{r}
param_f <- model_f$par
param_m <- model_m$par
sd_f <- sqrt(diag(solve(model_f$hessian)))
sd_m <- sqrt(diag(solve(model_m$hessian)))
model_result_f <- cbind(param_f, sd_f)
model_result_m <- cbind(param_m, sd_m)
rownames(model_result_f) <- c("L inf", "K", "t0")
rownames(model_result_m) <- c("L inf", "K", "t0")

kable(model_result_f, col.names = c("Estimated parameters", "Standard deviation"), caption = "Model for female fish body length", digits = 3)

kable(model_result_m, col.names = c("Estimated parameters", "Standard deviation"), caption = "Model for male fish body length", digits = 3)
```

Regression model for female fish: *Body length = `r round(param_f[1], 2)`(1 - exp(-`r round(param_f[2], 2)`(Age - `r round(param_f[3], 2)`)))*

Regression model for male fish: *Body length = `r round(param_m[1], 2)`(1 - exp(-`r round(param_m[2], 2)`(Age - `r round(param_m[3], 2)`)))*


### Visualize estimations of regression model
```{r}
#prepare data for plotting
dataset = data.frame(c(age, age[-c(12:13)]), c(length_f, length_m), c(VBGF(age, param_f[1], param_f[2], param_f[3]), VBGF(age[-c(12:13)], param_m[1], param_m[2], param_m[3])), c(rep("Female", 13), rep("Male", 11)))
colnames(dataset) <- c("Age", "Actual length", "Esimated length", "Sex")

#draw graph
ggplot(dataset, aes(x = Age)) +
    geom_point(aes(y = `Actual length`, col = Sex)) +
    geom_line(aes(y = `Esimated length`, col = Sex)) +
    scale_color_discrete(name = "Sex") +
    labs(title = "Regression model using von Bertalanffy growth function", y = "Fish length")
```