---
title: "HW #10 Computer Intensive Statistics in Ecology"
author: "Ä¬¥D©¶ Eric Su"
date: "2017/5/4"
output: html_document
---
[Click here for other works of the author on RPubs](http://rpubs.com/prorichter)

### Load packages
```{r message=FALSE}
library(ggbiplot)
library(vegan)
library(cluster)
```

## Q. Use the ¡§dominant¡¨ copepod species data (from HW1). Apply PCA or MDS using matrix algebra to extract major gradients of the dominant species. Make a bi-plot to show the relationships among species and sites. Then, check your results with those from build-in functions of PCA, MDS, and NMDS.

### Import data
Import copepod data and extract dominant species data. Also extract group from the result of hierarchical clustering in [previous homework](http://rpubs.com/prorichter/267782).
```{r}
#import copepod data
copdata = read.table("copepod_composition.txt", header = T)

#import species name
species_name <- read.table("copepodSPlist.txt", sep = "\t")

#assign copepod species name
rownames(copdata) <- as.character(unlist(species_name))

#convert species frequency into percentage
copdata = copdata / 100

#find dominant species
dom = apply(copdata >= 0.02, 2, which)
dom = sort(unique(unlist(dom)))
dom = t(copdata[dom, ])

#calculate distance matrix
dist_dom = vegdist(dom, method = "jaccard")

#perform hierarchical clustring
hc_dom = agnes(dist_dom, method = "ward")
group = cutree(hc_dom, 3)
group = factor(group, labels = c("Spring", "Summer", "Winter"))
```

### Conduct eigen analysis on dominant copepod data
```{r}
#eigen analysis on correlation matrix
dom_eigen <- eigen(cor(dom))

#extract pc loadings
loading = dom_eigen$vectors
rownames(loading) <- colnames(dom)

#calculate pc scores
score = scale(dom) %*% loading
```

### Determine number of PCs to keep
```{r}
var_expla = dom_eigen$values / sum(dom_eigen$values)
plot(var_expla, type = "b")
plot(cumsum(var_expla), type = "b")
```

It seems that the first 4 PCs explains most of the variance. 

### Create bi-plot to visualize result
```{r fig.width=15, fig.height=15}
biplot(cbind(score[,1], score[,2]), loading[, 1:34], ylabs = colnames(dom), xlab = paste("PC1(", 100 * round(var_expla[1], 2), "%)", sep = "" ), ylab = paste("PC2(", 100 * round(var_expla[2], 2), "%)", sep = "" ))
```

### Loading plots {.tabset}

Create loading plots to inspect copepods' effect on each PC.
```{r results = "asis", fig.width=15, fig.height=15}
for(i in 1:4){
    cat("\n")
    cat("#### ", "PC", i, "\n")
    pc_load = loading[, i]
    dotchart(pc_load[order(pc_load, decreasing = FALSE)], main = paste("Loading Plot for PC", i, sep = ""), xlab = "Variable Loadings", col = "blue")
    cat("\n")
}
```

### PCA
Conduct PCA using build-in function `prcomp`
```{r}
dom_pca <- prcomp(dom, scale. = T)
```

### Visulize PCA results using bi-plot {.tabset}

#### PCA bi-plot
```{r fig.width=15, fig.height=15}
biplot(dom_pca)
```

#### Compare with matrix algebra result
Since the +/- sign of the PCs is arbitrary, we can reverse them without distorting the results. This graph is made after reversing the +/- of eigen vectors.
```{r fig.width=15, fig.height=15}
biplot(cbind(-score[, 1], -score[, 2]), -loading[, 1:34], ylabs = colnames(dom), xlab = paste("PC1(", 100 * round(var_expla[1], 2), "%)", sep = "" ), ylab = paste("PC2(", 100 * round(var_expla[2], 2), "%)", sep = "" ))
```

###

The two bi-plots are very similar, indicating that we have done the eigen analysis correctly.

### Add group to graph
In this bi-plot, groups from hierarchical clustering are added and also a prettier graph is used
```{r  fig.width=15, fig.height=15}
ggbiplot(dom_pca, obs.scale = 1, var.scale = 1, ellipse = TRUE, groups = group,
         circle = TRUE) +
    scale_color_discrete(name = '') +
    theme(legend.direction = 'horizontal', legend.position = 'top') +
    theme_bw()
```

We can see that the first two principle components seperates summer from the other seasons but does not seperates spring and winter too well.

### Other unconstrained ordination methods {.tabset}

#### MDS
Conduct MDS using build-in function `cmdscale`
```{r warning=F}
dom_mds <- cmdscale(dist(dom), k = 4, eig = T)
dom_mds$points

#visulize result
ordiplot(dom_mds)
```


#### CA
Conduct CA using build-in function `cca`
```{r}
dom_ca <- cca(dom)
summary(dom_ca)

#Visulize CA result
ordiplot(dom_ca) 
```

#### DCA
Conduct DCA using build-in function `decorana`
```{r}
dom_dca = decorana(dom)
summary(dom_dca)

#Visulize DCA result
ordiplot(dom_dca, color = group) 
```

#### NMDS
Conduct NMDS using build-in function `metaMDS`
```{r}
dom_nmds = metaMDS(dom, trace = F, k = 4)
stressplot(dom_nmds)
dom_nmds$points
dom_nmds$species

#Visulize NMDS result
ordiplot(dom_nmds) 
```




